Monte Carlo solvers for sparse linear systems

### 2015-04-29 doc/ned2015/response_to_reviewer.txt

Response to reviewer's comments:

------------------------------------------------------------------------------
1. Since the underlying analysis seems to be graphical, they should use as
many graphs to explain what is happening.  Following words and moving
information is not so easy for a reader.

We agree that without visual aids the discussion is complex and is very much
based on the underlying graph of the matrix. We have added a figure (Figure 1)
in section 3.1 and discussion of the random walk in terms of the sparse graph
underneath the matrices that should help clarify the explanation, particularly
for the n_s and n_i parameters in our diffusion analysis. In addition, we have
expanded our description of the algorithm in section 2 and have provided more
definitions for clarity.

To improve clarity of the results, we also have re-plotted our results using a
different plotting tool to create higher quality images that should aid the
reader. Note that the y-axis in figure 2 has been changed to a logarithmic
scale and the random walk length and error measurements have been broken into
two different plots.

@@

Although we agree other matrix orderings could have been used, the results of
the paper are fundamentally based on the local stencil and independent the
overall global structure of the matrix. A block tridiagonal ordering could
reduce the bandwidth of the matrix and improve computational efficiency for a
matrix-vector multiply. However, the Monte Carlo algorithm would perform
identically in both cases given the same parallel decomposition of the
problem. The addition of Figure 1 in response to the first comment should also
help with this part of the analysis.

For example, given a banded ordering like that used in this work or the
suggested block tridiagonal ordering, the probability distribution functions
would be of the same size and have the same nearest-neighbor connections
because both use the same local stencil. Therefore, both cases will yield
random walks of the same time and space complexity. In addition, if both cases
had the same domain decomposition, random walks would be communicated from
domain to domain in an identical matter because the same local stencil was
used, therefore also resulting in the same parallel analysis.

------------------------------------------------------------------------------
9. Finally, there are no flux results with estimated errors.  We need results
since, after all, they are the bottom-line.

The purpose of the paper is to provide analytic relationships for the
domain-to-domain communication of random walks, and therefore a measure of
potential bandwidth and latency issues, when the Neumann-Ulam algorithm is
subject to domain decomposition. The domain decomposition itself does not
perturb the results of the stochastic algorithm (e.g. the flux in this
case). Rather, it simply provides a means to perform random walks via the
rules of the Neumann-Ulam game where the data for the random walks (weights
and probabilities) has been domain-decomposed.

Variance estimation, however, is a different story and domain decomposition
schemes to date trade inexact variance estimates for scalability. This is due
to the fact that the order of events must be maintained to achieve a correct
variance estimate and fully asynchronous algorithms do not, in general, permit
this. The literature indicates that this order of operations is less
significant for the correctness of the flux estimate with differences in
computations induced by round-off error in floating point
computations. Furthermore, the accuracy and convergence properties of the
Monte Carlo method have been well documented in the references cited by the
paper.

### 2015-09-16 MAX_reports/MC_solvers_article/MC_solvers.tex
Restructuring paper into smaller number of sections.

\section{Stochastic setting}

We start our treatise by introducing the direct methods.\newline

Assume to start from a linear system such as 
\begin{equation}
A \mathbf{x}=\mathbf{b},
\label{linsys}
\end{equation}
where $A\in \mathbb{R}^{n\times n}$, $\mathbf{x}$, $\mathbf{b} \in 
\mathbb{R}^n$.

@@

\subsection{Forward Method}
The goal is to evaluate a functional such as 
\[
J(\mathbf{x})=(\mathbf{h},\mathbf{x})=\sum_{i=1}^n h_i x_i.
\]
where $\mathbf{h}\in \mathbb{R}^n$ is the Riesz representative in 
$\mathbb{R}^n$ of the functional $J$.
We can use it to build the initial probability $\tilde{p}: 
S\rightarrow [0,1]$ of the random walk such that
\[
\tilde{p}(k_0=i)=\tilde{p}_{k_0}=\frac{\lvert h_i\rvert}{\sum_{i=1}^n \lvert 
h_i\rvert}.
\]
It is important to stress out that the role of vector $\mathbf{h}$ is 
restricted to the construction of the initial probability. Once that is done, 
it is 
not used anymore in the definition of the stochastic process. 
A possible choice for the transition probability $P$ can be
\[
p(k_i=j \;\lvert\;k_{i-1}=i )=P_{i,j}=\frac{\lvert H_{i,j}\rvert}{\sum_{k=1}^n 
\lvert H_{i,k}\rvert}.
\]
where $\tilde{p}(\cdot,i):S\rightarrow [0,1]$ $\forall i\in S$.
A related sequence of random variables $w_{i,j}$ can be defined 
such that 
\[
w_{i,j}=\frac{H_{i,j}}{P_{i,j}}.
\]
The probability distribution of the random variables $w_{i,j}$ is represented 
by the transition matrix that rules the stochastic process. The $w_{i,j}$'s 
just introduced can be used to build one more sequence 
of random variables. 
At first we introduce quantities $W_j$ 
\[
W_{0}=\frac{h_{k_0}}{\tilde{p}_{k_0}}, \quad W_j=W_{j-1} w_{i,j}, \quad 
j=1,\cdots, i.
\]
By defining 
\[
X(\nu)=\sum_{m=0}^k W_m f_{i_m}
\]
as the random variable associated with a specific permutation $\nu$, we can 
define the estimator $\theta_i$ such as
\[
\theta=E[X]=\sum_{\nu}P_{\nu}X(\nu).
\]
The integer $n$ represents the size of the solution vector to (\ref{linsys}) 
and 
the
index 
$i$ 
is referred to the component of the solution vector we want to compute. 
$P_{\nu}$ is the probability associated with a specific permutation of the 
random walk.  
It can be proved that 
\[
E[W_i f_{k_i}]=(\mathbf{h},H^i\mathbf{f}), \quad i=0,1,2,\cdots
\]
and 
\[
\theta_i=E\bigg[\sum_{i=0}^\infty W_i f_{k_i}\bigg]=(\mathbf{h},\mathbf{x}).
\]

@@

\subsection{Adjoint Method}
A second Monte Carlo method can be derived by considering the linear system 
adjoint to (\ref{linsys_prec})
\begin{equation}
(P^{-1}A)^T\mathbf{y}=\mathbf{d},
\label{adj}
\end{equation}
where $\mathbf{y}$ and $\mathbf{d}$ are the adjoint solution and source term.

@@

\section{Numerical interpretation of the statistical requirements}
The requests imposed on the expected value and second moment of the estimators 
can be reformulated in a purely deterministic setting by looking at the 
spectral radii of some properly defined matrices.

@@

\section{Necessary conditions, sufficient conditions and choice of the 
transition matrix}

@@

\section{Construction of the transition probability}
The way the transition probability is defined is extremely important. In fact 
in some circumstances it might determine whether the stochastic scheme 
converges or not. Two very simple ways of defining it are listed here 
below. These are the approaches adopted typically in literature.

\subsection*{Uniform probabilities}
With this approach transition matrix $P$ is such that all the non-zero elements 
in each row have equal probability of occurring. 
\[
\text{Forward}:\;P_{i,j}=
\begin{cases}
0 \quad \quad \quad \qquad \qquad \qquad \text{if}\quad H_{i,j}=0 \\
\frac{1}{\#(\text{non-zeros in the row)}} \quad \text{if} \quad H_{i,j}\ne 0
\end{cases}\quad 
\]
\[
\text{Adjoint}:\;P_{i,j}=
\begin{cases}
0 \quad \quad \quad \qquad \qquad \qquad \text{if}\quad H_{j,i}=0 \\
\frac{1}{\#(\text{non-zeros in the column)}} \quad \text{if} \quad H_{j,i}\ne 0
\end{cases}
\]
The Monte Carlo approach resorting to this definition of the transition matrix, 
in accordance to \cite{AADBTW2005}, is called \textit{Uniform Monte Carlo} (UM).

\subsection*{Weighted probabilities}
An alternative way of defining the transition matrix consists in attributing 
probabilities to all the entries of the transition probability accordingly 
to the magnitude of the elements. For instance we may employ the following 
definition
\[
\text{Forward}: \;
p(k_i=j \;\lvert\;k_{i-1}=i )=P_{i,j}=\frac{\lvert 
H_{i,j}\rvert^p}{\sum_{k=1}^n 
\lvert H_{i,k}\rvert^p}
\]

@@

\subsection*{Examples of matrices for which $\rho(H)<1$ and $\rho(\hat{H})>1$}
By focusing on the iteration matrix related to the fixed point scheme, we 
provide some examples of matrices for which the traditional condition 
$\rho(H)<1$ is not mirrored by the equally necessary condition 
$\rho(\hat{H})<1$.
The Almost Optimal transition probability is employed in order to achieve 
optimal upper bounds for the spectral radius of $\hat{H}$. The iteration matrix 
is 
obtained by applying a left diagonal preconditioner to the original system 
\ref{linsys}.

The matrices we consider are taken from the Matrix Market repository. The first 
one is "JPWH\_991". As it can be noticed from Table \ref{table_nonconv}, the 
Forward MC guarantees convergence since both the the conditions on $H$ and 
$\hat{H}$ 
are verified. The same does not hold for the Adjoint method, since the condition 
on $\hat{H}$ does not hold anymore.

Another example provided is the matrix "FS\_680\_1". In this case both the 
Forward and the Adjoint methods do not converge, since the spectral radius of 
$\hat{H}$ is bigger than 1 for both the methods.

\begin{table}[!h]
\centering
\begin{tabular}{|c|c|c|c|c|}
\hline
\textbf{Type of problem} & \textbf{d.o.f.'s} &
$\rho(H)$&Forward $\rho(\hat{H})$&Adjoint $\rho(\hat{H})$\\
\hline
JPWH\_991 & 991 & $0.9797$ & $0.9797$ & $1.0505$\\ 
\hline
FS\_680\_1 & 680 & $0.9697$ & $1.2554$ & $3.5984$\\
\hline
\end{tabular}
\caption{Examples of matrices for which $\rho(\hat{H})<1$ is violated.} 
\label{table_nonconv}
\end{table}

\section{Adaptive approaches}

\subsection{An adaptive cut-off for the length of random walks}

\subsection{An a posteriori variance-based adaptivity}

\section{Sets of matrices for which the convergence of the MC linear solver is 
guaranteed}

@@

 \subsection{Strictly diagonally dominant matrices}
 One of the these sets is 
represented by strictly diagonally dominant matrices.

@@

 \subsection{Generalized diagonally dominant matrices (GDDM)}

@@

 \subsection{"Block-diagonally dominant"-like matrices}
 In this section we wonder in which situations a block diagonal preconditioning 
might come up with a convergent Monte Carlo linear solver. 
By mimicking the computations we already showed previously, the iteration 
matrix $H\in\mathbb{R}^{n\times n}$ resulting from a block diagonal 
preconditioning takes the form

@@

\subsubsection*{Forward method}
By assuming that all $n_i$'s have the same value we may define
\[
 m=n_i=\text{size of a block}=\frac{n}{p}.
\]

@@

\subsubsection*{Adjoint method} 
Analogously to the Forward method, if we define 
\[
(\hat{H})^T_{i,j} = \lvert H^T_{i,j}\rvert\bigg(\sum_{k=1}^n\lvert 
H^T_{ik}\rvert\bigg)
\]
we can formulate a sufficient condition for the convergence of the Adjoint 
Monte 
Carlo method by introducing a matrix $\tilde{H}$ which in this case is such that
\[
  \tilde{H}=\begin{bmatrix}0 & \lVert [A_{11}]^{-1}A_{12}\rVert_{1} & \cdots 
&
\cdots & \lVert [A_{11}]^{-1}A_{1p}\rVert_{1} \\
\lVert [A_{22}]^{-1}A_{21}\rVert_{1} & 0 & \lVert 
[A_{22}]^{-1}A_{23}\rVert_{1} & 
\cdots & \lVert [A_{22}]^{-1}A_{2p}\rVert_{1} \\
\vdots & \vdots & \ddots & \vdots & \vdots\\
\vdots & \vdots & \vdots &\ddots & \vdots \\
\lVert [A_{pp}]^{-1}A_{p1}\rVert_{1} &  \cdots & \cdots& 
\lVert [A_{pp}]^{-1}A_{p,p-1}\rVert_{1} & 0
\end{bmatrix}.
\]
The sufficient condition assumes the form
\begin{equation}
 \lVert \tilde{H} \rVert_{1}<1 \Rightarrow \lVert \hat{H} \rVert_{1}<1.
\end{equation}

\section{Linear solvers with hybrid schemes}

\subsection{Preconditioning techniques}
In this section we focus on the pursuit of preconditioners that facilitate the 
convergence of the stochastic schemes introduced so far. A priori, all of the 
following techniques are good candidates to improve the numerical setting. 
However issues about the fill-in effect and the verification of convergence 
conditions limit their employment for our purposes.

@@

\section{Numerical results}

All the numerical results shown in the following sections are computed with 
\texttt{Matlab} on a standard laptop. The simulations are accomplished in a 
serial mode.
In the 
first subsection the standard Monte Carlo solver is employed, while in the 
subsequents ones results are obtained with the MCSA scheme.

\subsection{Check on the effectiveness of the adaptive criteria}
In this subsection we show some numerical tests to validate the utility of the 
adaptive approaches for the selection of the number of permutations to employ. 
For this purpose, we restrict our attention to standard Monte Carlo linear 
solvers.

@@

\subsection{A comparison between different preconditioners and their effect 
on $\rho(H)$ and $\rho(\hat{H})$}

@@

\subsection{The Poisson equation solved with MCSA}
In the current subsection a Poisson problem is taken into account:
\begin{equation}
\begin{cases}
 -\Delta u = f \quad \text{in}\quad \Omega \\
 u=0\quad \text{on} \quad \partial\Omega
 \end{cases}
\end{equation}

@@

\subsection{A diffusion reaction problem solved with MCSA}
By modifying the equation of the previous test case, we consider now 
an advection reaction problem

@@

\subsection{A parabolic problem solved with MCSA}
In this subsection we restrict our treatise to the analysis of parabolic 
partial 
differential problems of the form 
\begin{equation}
\begin{cases}
 \frac{\partial u }{\partial t} + \mathcal{L}(u)=f \quad \text{in} \Omega \\
\text{B.C.} \quad \text{on}\quad \delta \Omega.
\end{cases}
\label{parabolic}
\end{equation}

@@

Other preconditioners 
have 
been 
tested besides the sparse approximate inverse preconditioner, such as block 
diagonal 
and incomplete LU factorization. However the fill-in effect introduced by the 
application of 
these preconditioners compromises the sparsity of the linear system.

\section{Conclusions and future developments}
The work presented in this paper has been accomplished to provide an overview 
about MC linear solvers. In particular the goal was to combine the main 
theoretical results found in literature with additional novel contributions on 
both the theoretical and empirical sides. This made it possible to understand 
the actual limitations of the presented algorithms.

>>>>

\section{Stochastic Linear Solvers}

We consider the solution of systems of linear equations of the form
\begin{equation}
A \mathbf{x}=\mathbf{b},
\label{linsys}
\end{equation}
where $A\in \mathbb{R}^{n\times n}$, $\mathbf{x}$, $\mathbf{b} \in
\mathbb{R}^n$.

@@

\subsection{Forward Method}

The goal is to evaluate a functional such as
\[
J(\mathbf{x})=(\mathbf{h},\mathbf{x})=\sum_{i=1}^n h_i x_i.
\]
where $\mathbf{h}\in \mathbb{R}^n$ is the Riesz representative in
$\mathbb{R}^n$ of the functional $J$.
We can use it to build the initial probability $\tilde{p}:
S\rightarrow [0,1]$ of the random walk such that
\[
\tilde{p}(k_0=i)=\tilde{p}_{k_0}=\frac{\lvert h_i\rvert}{\sum_{i=1}^n \lvert
h_i\rvert}.
\]
It is important to stress out that the role of vector $\mathbf{h}$ is
restricted to the construction of the initial probability. Once that is done,
it is
not used anymore in the definition of the stochastic process.
A possible choice for the transition probability $P$ can be
\[
p(k_i=j \;\lvert\;k_{i-1}=i )=P_{i,j}=\frac{\lvert H_{i,j}\rvert}{\sum_{k=1}^n
\lvert H_{i,k}\rvert}.
\]
where $\tilde{p}(\cdot,i):S\rightarrow [0,1]$ $\forall i\in S$.
A related sequence of random variables $w_{i,j}$ can be defined
such that
\[
w_{i,j}=\frac{H_{i,j}}{P_{i,j}}.
\]
The probability distribution of the random variables $w_{i,j}$ is represented
by the transition matrix that rules the stochastic process. The $w_{i,j}$'s
just introduced can be used to build one more sequence
of random variables.
At first we introduce quantities $W_j$
\[
W_{0}=\frac{h_{k_0}}{\tilde{p}_{k_0}}, \quad W_j=W_{j-1} w_{i,j}, \quad
j=1,\cdots, i.
\]
By defining
\[
X(\nu)=\sum_{m=0}^k W_m f_{i_m}
\]
as the random variable associated with a specific permutation $\nu$, we can
define the estimator $\theta_i$ such as
\[
\theta=E[X]=\sum_{\nu}P_{\nu}X(\nu).
\]
The integer $n$ represents the size of the solution vector to (\ref{linsys})
and
the
index
$i$
is referred to the component of the solution vector we want to compute.
$P_{\nu}$ is the probability associated with a specific permutation of the
random walk.
It can be proved that
\[
E[W_i f_{k_i}]=(\mathbf{h},H^i\mathbf{f}), \quad i=0,1,2,\cdots
\]
and
\[
\theta_i=E\bigg[\sum_{i=0}^\infty W_i f_{k_i}\bigg]=(\mathbf{h},\mathbf{x}).
\]

@@

\subsection{Adjoint Method}

A second Monte Carlo method can be derived by considering the linear system
adjoint to (\ref{linsys_prec})
\begin{equation}
(P^{-1}A)^T\mathbf{y}=\mathbf{d},
\label{adj}
\end{equation}
where $\mathbf{y}$ and $\mathbf{d}$ are the adjoint solution and source term.

@@

\subsection{Hybrid Stochastic/Deterministic Methods}

\subsubsection{Sequential Monte Carlo}

\subsubsection{Monte Carlo Synthetic Acceleration}

\section{Convergence Behavior of Stochastic Methods}

The requests imposed on the expected value and second moment of the estimators
can be reformulated in a purely deterministic setting by looking at the
spectral radii of some properly defined matrices.

@@

\subsection{Necessary and Sufficient Conditions}

@@

\subsubsection{Forward MC}

\subsubsection{Adjoint MC with Collision Estimator}

\subsubsection{Adjoint MC with Expected-Value Estimator}

\subsection{Construction of Transition Probabilities}

The way the transition probability is defined is extremely important. In fact
in some circumstances it might determine whether the stochastic scheme
converges or not. Two very simple ways of defining it are listed here
below. These are the approaches adopted typically in literature.

\subsubsection{Uniform probabilities}
With this approach transition matrix $P$ is such that all the non-zero elements
in each row have equal probability of occurring.
\[
\text{Forward}:\;P_{i,j}=
\begin{cases}
0 \quad \quad \quad \qquad \qquad \qquad \text{if}\quad H_{i,j}=0 \\
\frac{1}{\#(\text{non-zeros in the row)}} \quad \text{if} \quad H_{i,j}\ne 0
\end{cases}\quad
\]
\[
\text{Adjoint}:\;P_{i,j}=
\begin{cases}
0 \quad \quad \quad \qquad \qquad \qquad \text{if}\quad H_{j,i}=0 \\
\frac{1}{\#(\text{non-zeros in the column)}} \quad \text{if} \quad H_{j,i}\ne 0
\end{cases}
\]
The Monte Carlo approach resorting to this definition of the transition matrix,
in accordance to \cite{AADBTW2005}, is called \textit{Uniform Monte Carlo} (UM).

\subsubsection{Weighted probabilities}
An alternative way of defining the transition matrix consists in attributing
probabilities to all the entries of the transition probability accordingly
to the magnitude of the elements. For instance we may employ the following
definition
\[
\text{Forward}: \;
p(k_i=j \;\lvert\;k_{i-1}=i )=P_{i,j}=\frac{\lvert
H_{i,j}\rvert^p}{\sum_{k=1}^n
\lvert H_{i,k}\rvert^p}
\]

@@

\subsection{Classes of Matrices with Guaranteed Convergence}

@@

\subsubsection{Strictly Diagonally Dominant}

 One of the these sets is
represented by strictly diagonally dominant matrices.

@@

\subsubsection{Generalized Diagonally Dominant}

@@

\subsubsection{Block Diagonally Dominant}

 In this section we wonder in which situations a block diagonal preconditioning
might come up with a convergent Monte Carlo linear solver.
By mimicking the computations we already showed previously, the iteration
matrix $H\in\mathbb{R}^{n\times n}$ resulting from a block diagonal
preconditioning takes the form

@@

\paragraph{Forward method}
By assuming that all $n_i$'s have the same value we may define
\[
 m=n_i=\text{size of a block}=\frac{n}{p}.
\]

@@

\paragraph{Adjoint method}
Analogously to the Forward method, if we define
\[
(\hat{H})^T_{i,j} = \lvert H^T_{i,j}\rvert\bigg(\sum_{k=1}^n\lvert
H^T_{ik}\rvert\bigg)
\]
we can formulate a sufficient condition for the convergence of the Adjoint
Monte
Carlo method by introducing a matrix $\tilde{H}$ which in this case is such that
\[
  \tilde{H}=\begin{bmatrix}0 & \lVert [A_{11}]^{-1}A_{12}\rVert_{1} & \cdots
&
\cdots & \lVert [A_{11}]^{-1}A_{1p}\rVert_{1} \\
\lVert [A_{22}]^{-1}A_{21}\rVert_{1} & 0 & \lVert
[A_{22}]^{-1}A_{23}\rVert_{1} &
\cdots & \lVert [A_{22}]^{-1}A_{2p}\rVert_{1} \\
\vdots & \vdots & \ddots & \vdots & \vdots\\
\vdots & \vdots & \vdots &\ddots & \vdots \\
\lVert [A_{pp}]^{-1}A_{p1}\rVert_{1} &  \cdots & \cdots&
\lVert [A_{pp}]^{-1}A_{p,p-1}\rVert_{1} & 0
\end{bmatrix}.
\]
The sufficient condition assumes the form
\begin{equation}
 \lVert \tilde{H} \rVert_{1}<1 \Rightarrow \lVert \hat{H} \rVert_{1}<1.
\end{equation}

\subsection{Adaptive Methods}

\subsubsection{History Length}

\subsubsection{Number of Random Walks}

\subsection{Preconditioning}

In this section we focus on the pursuit of preconditioners that facilitate the
convergence of the stochastic schemes introduced so far. A priori, all of the
following techniques are good candidates to improve the numerical setting.
However issues about the fill-in effect and the verification of convergence
conditions limit their employment for our purposes.

@@

\section{Numerical Results}

\subsection{Adaptive Methods}

In this subsection we show some numerical tests to validate the utility of the
adaptive approaches for the selection of the number of permutations to employ.
For this purpose, we restrict our attention to standard Monte Carlo linear
solvers.

@@

\subsection{Preconditioning Approaches}

@@

\subsection{Hybrid Methods}

\subsubsection{Poisson Problem}

In the current subsection a Poisson problem is taken into account:
\begin{equation}
\begin{cases}
 -\Delta u = f \quad \text{in}\quad \Omega \\
 u=0\quad \text{on} \quad \partial\Omega
 \end{cases}
\end{equation}

@@

\subsubsection{Reaction-Diffusion Problem}

By modifying the equation of the previous test case, we consider now
an advection reaction problem

@@

\subsubsection{Parabolic Problem}

In this subsection we restrict our treatise to the analysis of parabolic
partial
differential problems of the form
\begin{equation}
\begin{cases}
 \frac{\partial u }{\partial t} + \mathcal{L}(u)=f \quad \text{in} \Omega \\
\text{B.C.} \quad \text{on}\quad \delta \Omega.
\end{cases}
\label{parabolic}
\end{equation}

@@

\section{Conclusions and Future Work}

The work presented in this paper has been accomplished to provide an overview
about MC linear solvers. In particular the goal was to combine the main
theoretical results found in literature with additional novel contributions on
both the theoretical and empirical sides. This made it possible to understand
the actual limitations of the presented algorithms.
